SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-05 design_arch.tex
Some small typos

As mentioned before, ultimately it might be benefitial to allow a
prefetcher to speculate as deeply as possible. Often, some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below a point, in which performance improvements are observed, due to the
increase of inaccurate prefetches.  However, in order to allow deep
speculation in the prefetcher, inaccurate prefetches must be
filtered out.  We propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones. The Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, as illuastrated
in figure ~\ref{fig:PPF_Hierarchy}, to prevent
not-useful prefetches from polluting the higher levels of the memory hierarchy.

As discussed in Section~\ref{sec:Background-Perceptron}, perceptron learning
is a light-weight mechanism to pull together disparate forms of information
and synthesize a decision from them.  Our work considers a number of
features corresponding to a prefetch, such as speculation
depth, page address and offset, and use this information as input to our perceptron-based filter in order to predict the usefulness of a prefetch.
In the remainder of this section, we discuss our proposed technique PPF.
It enhances a base prefetcher, by filtering out not-useful prefetches. 
PPF is a generalized prefetch filtering mechanism that may be adapted 
to any prefetcher with appropriate modifications.

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as the
steps followed to filter out not-useful prefetches.
The filter is organized as a set of tables, where each entry in a tables holds 
a \textit{weight}. For a configuration of PPF using $N$ number of features, 
$N$ different tables of weights are needed. Each feature is used to index 
a distinct table. The number of entries of each table varies according
to their corresponding feature, hence, different
number of bits are needed to index different tables.
Each weight is a 5-bit saturating counter ranging from -16 to +15. 
We found the having 5-bit weights provide a good trade-off 
between accuracy and area. A detail explanation of the storage overhead of
PPF can be found in Section~\ref{sec:Method-Overheads}.

@@

%ETG:is this like an explanation of how prefetching happens?

The base prefetcher is triggered on every demand access to the L2.  
As this point, it has the opportunity to trigger a prefetch  -- if it 
decides to do so, it will also need to decide how many cache lines
to prefetch.  These cache lines can be either placed in the L2 or
L3 according to the confidence of the prefetching mechanism.
Once the base prefetcher is triggered, the suggested
prefetch candidates are feeded to the perceptron filter, 
in order to determine the usefulnes of these prefetches.  
The filter makes the decision to issue a prefetch request based on 
the suggestions from the base prefetcher.
In order to make such decision, each feature corresponding to a 
suggested prefetch is used to index a table, 
all corresponding weights are summed. 
The sum denotes the confidence value for the suggested prefetch, 
and is thresholded against two diffent magnitudes:  
PERC\_THRESHOLD\_HI and PERC\_ THRESHOLD\_ LO.

Prefetches which sum is above PERC\_THRESHOLD\_HI
are placed into the L2 cache, the higher confidence value
hints the prefetch would be useful and should be prioritized.
A prefetch for which its corresponding features result 
in a confidence value between  
PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI is allocated in the larger 
last-level cache, the filter is moderately confident of the future reuse of the 
cache line, but not enough to possible pollute a significantly smaller L2. 
Suggested prefetches which features yield a confident value lower than 
PERC\_ THRESHOLD\_ LO are not prefetched, a low confidence value 
represents that the filter has learn that similar set of features
are associated with not-useful prefetches. \newline

@@

The address from the demand request that trigered training is lookup in 
both tables. If the address resides in the prefetch table, and is marked 
as valid, this hints the previous prediction was correct and this is a useful
prefetch. We compute the sum of the corresponding weights and
compare the sum to an specific threshold, if the sum falls below it, 
training must happen and the corresponding weights are adjusted accordingly.
These thresholds are introduced inorder to avoid overtraining, 
which helps the filter adapt quickly to changes in memory behavior. 
These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

On an eviction of a cache block, we look up the corresponding address
in the prefetch table. If there is a valid entry with this address, 
this mean the filter made a misprediction. The block was allocated in the
L2 with a prefetch request that the filter should have categorized 
not-reuse prefetch. In this scenario, the corresponding features of the
prefetch request that cause these block to be placed in the cache are used
to index the tables of weights, and adjust weights accordingly.

%wait at the same time or no?
Paralell to accessing the prefetch table, on a demand access, 
the reject table is accessed. Before
the demand access triggers the next set of prefetches, the reject table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the base prefetcher, but wrongly rejected by the
perceptron filter. The perceptron filter learns from this and makes use of the
correspoding features associated to the original prefetch request, which are
stored in the reject table, to index the weights tables and adjust the weigths
accordingly.

Prior work lacks the knowledge of which cache lines should have 
been prefetched, but were not. The implementation of the reject table,
allows us to have this information, which can be use to futher optimize
our prefetching mechanism.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be adapted
to be used over any base prefetcher for increased prefetch accuracy.  In
general, only two hooks are required between PPF and the baseline prefetcher.
The first is to make sure that all the prefetch candidates of the prefetcher
pass through the perceptron filter and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

>>>>

As mentioned before, ultimately it might be beneficial to allow a
prefetcher to speculate as deeply as possible. Often, some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below a point, after which the performance degrades, due to the
increase of inaccurate prefetches.  However, in order to allow deep
speculation in the prefetcher, inaccurate prefetches must be
filtered out.  We propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones. The Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, as illustrated
in figure ~\ref{fig:PPF_Hierarchy}, to prevent
not-useful prefetches from polluting the higher levels of the memory hierarchy.

As discussed in Section~\ref{sec:Background-Perceptron}, perceptron learning
is a light-weight mechanism to pull together disparate forms of information
and synthesize a decision from them.  Our work considers a number of
features corresponding to a prefetch, such as speculation
depth, page address and offset, and uses this information as the inputs to our 
perceptron-based filter, in order to predict the usefulness of a prefetch.
In the remainder of this section, we discuss our proposed technique -- PPF.
It enhances a base prefetcher, by filtering out not-useful prefetches. 
PPF is a generalized prefetch filtering mechanism that may be adapted 
to any prefetcher with appropriate modifications.

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as the
steps followed to filter out not-useful prefetches.
The filter is organized as a set of tables, where each entry in the tables holds 
a \textit{weight}. For a configuration of PPF using $N$ number of features, 
$N$ different tables of weights are needed. Each feature is used to index 
a distinct table. The number of entries of each table varies according
to their corresponding feature, hence, different
number of bits are needed to index different tables.
Each weight is a 5-bit saturating counter ranging from -16 to +15. 
We found that having 5-bit weights provide a good trade-off 
between accuracy and area. A detail explanation of the storage overhead of
PPF can be found in Section~\ref{sec:Method-Overheads}.

@@

%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.

The base prefetcher is triggered on every demand access to the L2 Cache.  
As this point, it has the opportunity to trigger a prefetch  -- if it 
decides to do so, it will also need to decide how many cache lines
to prefetch.  These cache lines can be either placed in the L2 Cache or
the L3 Cache, according to the confidence of the prefetching mechanism.
Once the base prefetcher is triggered, the suggested
prefetch candidates are fed to the perceptron filter, 
in order to determine the usefulness of these prefetches.  
The filter makes the decision to issue the prefetch
suggestions from the base prefetcher.
In order to make such decision, each feature corresponding to a 
suggested prefetch is used to index a table, 
all the corresponding weights are summed. 
The sum denotes the confidence value for the suggested prefetch, 
and is thresholded against two different magnitudes:  
PERC\_THRESHOLD\_HI and PERC\_ THRESHOLD\_ LO.

Prefetches which sum is above PERC\_THRESHOLD\_HI
are placed into the L2 cache, the higher confidence value
hints the prefetch would be useful and should be prioritized.
A prefetch for which its corresponding features result 
in a confidence value between  
PERC\_THRESHOLD\_LO and PERC\_THRESHOLD\_HI is allocated in the larger 
last-level cache, as the filter is moderately confident of the future reuse of the 
cache line, but not enough to possible pollute a significantly smaller L2. 
Suggested prefetches for which the features lead to a confidence value lower than 
PERC\_THRESHOLD\_LO are not prefetched, as the low confidence value 
represents that the perceptron has learn that similar set of features
are associated with non-useful prefetches. \newline

@@

The address from the demand request that triggered the training is lookup in 
both tables. If the address resides in the prefetch table, and is marked 
as valid, this hints the previous prediction was correct and this is a useful
prefetch. We compute the sum of the corresponding weights and
compare the sum to an specific threshold, if the sum falls below it, 
training must happen and the corresponding weights are adjusted accordingly.
These thresholds are introduced in order to avoid over-training, 
which helps the filter adapt quickly to changes in memory behavior. 
These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

On an eviction of a cache block, we look up the corresponding address
in the prefetch table. If there is a valid entry with this address, 
this mean the filter made a misprediction. The block was allocated in the
L2 with a prefetch request that the filter should have categorized 
as a useless prefetch. In this scenario, the corresponding features of the
prefetch request that cause these block to be placed in the cache are used
to index the tables of weights, and those weights are adjusted accordingly.

%wait at the same time or no?
%Yes, and they are mutex cases

Parallel to accessing the prefetch table, on a demand access, 
the reject table is accessed. Before
the demand access triggers the next set of prefetches, the reject table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the base prefetcher, but wrongly rejected by the
perceptron filter. The perceptron filter learns from this and makes use of the
corresponding features associated to the original prefetch request, which are
stored in the reject table, to index the weights tables and adjust the weights
accordingly.

Prior work in this domain lacks the knowledge of which cache lines should have 
been prefetched, but were not. The implementation of the reject table,
allows us to have this information, which can be use to further optimize
our prefetching mechanism.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be adapted
to be used over any base prefetcher for increased prefetch accuracy. Only 
two hooks are required between PPF and the baseline prefetcher.
The first is to make sure that all the prefetch candidates of the prefetcher
pass through the perceptron filter and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.
